### 2021-08-31 response_to_review.Rmd

Thank you for the detailed and helpful reviews. These are the ways that we have changed the article in response to the comments.

# Co-editor

- I am fond of this reference, but perhaps Debby's contains it:

Salch, J.D. and D.W. Scott, (1997),
``Data Exploration with the Density Grand Tour,''
 Statistical Graphics and Computing Newsletter,
ASA, Alexandria, VA, 8: 7-11, April, 1997. https://community.amstat.org/jointscsg-section/newsletter

    - **Added**

@@

# Reviewer: 1

- Overall, there is nothing wrong in the article, but lots of language issues, problems with references, and a few aspects of tours (including software and references) that have not been explored in detail. Please see the attached markup for details.

    - **We are extremely grateful for the detailed read and fixes!**

@@

- p5 l49 "projections"

    - **Sentence changed**

- p5 l50 cite 1 or 2 references for projection pursuit, including the WIREs article - see
https://wires.onlinelibrary.wiley.com/doi/abs/10.1002/wics.23

    - **Introduced in section 1 where PCA and LDA is first mentioned**

@@

- p7 l37 be specific: "in Section xxx"

    - **done**

- p7 l39 same: "in Section yyy"

    - **Paragraph changed and it doesn't fit to forward reference here**

@@

- I am curious: Why do you not cite https://doi.org/10.1051/epjconf/202024506018

    - **This was a conference presentation and summarised work in other papers. It looks like a paper, which is the physics approach, but it cannot be considered a paper in its own right.**

### 2021-08-31 paper.Rmd
done to page 7

While tours are invaluable for assessing the geometry of data, they are by no means the only technique available for visualizing structure in high dimensional data. Very similar to a tour was the Andrews plot [@andrews] which uses a fourier series to create a smooth series of 1-D projections. But as @wegman-solka point out this is a pseudo-grand tour because it is not space-filling. An early technique proposed for assessing pairwise relationships between variables is the scatterplot matrix (SPLOM) [@Tukey1981-ov; @Chambers1983-dh; @Tukey1983-fj; @Carr1984-xw; @Becker1987-dw]. The SPLOM allows the viewer to assess correlation structure but does not scale to large numbers of variables. Similarly, parallel coordinates plots (PCP) can be used to explore correlation and collinearity [@Inselberg1985-kf; @Wegman1990-dy]. By placing multiple variables side-by-side in a PCP higher order, structure like clustering or lower dimensional embeddings are revealed. However the ordering of variables along the axis changes what can be learned, and @hurley-oldford suggest repeating variables to alleviate this. Another display that relies on variable ordering is the heatmap, which is widely used to visualize cluster structure in bioinformatics. @Wilkinson2009-pj provides a comprehensive history of this display in the social and natural sciences.

Recently, some review papers have summarized different aspects of tour methods. @Moustafa2009 explains the relationship between a grand tour and the classical multivariate plot, the Andrews curve. @Moustafa2010 describes a computational shortcut for a tour. @Martinez2013 explains the image grand tour.

@@

The recently developed slice tour [@Laa2020-js] can be applied to any of the above tour types. It displays a slice through the orthogonal space, as opposed to a projection, and is explained in more details later.

@@

The smooth progression in the tour path is due to geodesic interpolation between the target planes. This takes into consideration two important aspects: (1) maintains the orthonormality of the projection bases, (2) contains all the rotation to be between planes, not a particular basis in any plane. The first is clearly important because it ensures that we a looking at low-dimensional projection of the data always. The latter is harder to explain, but really important from a visual perspective. It stops any within-plane spin, and could be considered stabilizing the view. More details explanation of this can be found in @Buja2005-cx.

@@

Various displays are available to show projections in 1D, 2D and higher dimensions. A 1D projection displays the data in a histogram analogous to a shadow puppet being projected onto a wall, while 2D projections are displayed as scatterplots. Higher dimensional projections can be shown with multivariate displays like SPLOMs or parallel coordinates. Figure \ref{fig:display} shows a variety of displays for projections and these displays are useful to show the non-normal distributions in the projected data, clustering structure, and multivariate relationship between variables.

@@

## Slices, sections and projections

@@

There has been a long history of software including tour methods, or something similar. These have laid the foundation for the current tools. The American Statistical Association Statistical Graphics Video Library web site [@ASASGVL] hosts videos documenting the history. It is fabulous watching the videos titled "Multidimensional Scaling" (Kruskal, 1962), "Real-Time Rotation" (Chang, 1970) and "Prim-9" (Tukey, 1973) show some preliminary methods leading to the development of the grand tour. The video titled "Use of the Grand Tour in Remote Sensing" (McDonald and Willis, 1987) is the first to show a tour, and the video titled "Dataviewer: A Program for Looking at Data in Several Dimensions" (Buja and Tukey, 1987) demonstrates the tour as part of a larger data analysis system. The videos titled "XGobi: Dynamic Graphics for Data Analysis" (Swayne et al, 1991) and "Grand Tour and Projection Pursuit" (Cook et al, 1993) show the tour tools in the XGobi software system. Some work on tours, for example @Wegman92 and @Tierney1991 is not documented in the video collection.

>>>>

While tours are invaluable for assessing the geometry of data, they are by no means the only technique available for visualizing structure in high dimensional data. Very similar to a tour was the Andrews plot [@andrews] which uses a fourier series to create a smooth series of 1-D projections. But as @wegman-solka point out this is a pseudo-grand tour because it is not space-filling. @Moustafa2009 explains the relationship between a grand tour and the classical multivariate plot, the Andrews curve. An early technique proposed for assessing pairwise relationships between variables is the scatterplot matrix (SPLOM) [@Tukey1981-ov; @Chambers1983-dh; @Tukey1983-fj; @Carr1984-xw; @Becker1987-dw]. The SPLOM allows the viewer to assess correlation structure but does not scale to large numbers of variables. Similarly, parallel coordinates plots (PCP) can be used to explore correlation and collinearity [@Inselberg1985-kf; @Wegman1990-dy]. By placing multiple variables side-by-side in a PCP higher order, structure like clustering or lower dimensional embeddings are revealed. However the ordering of variables along the axis changes what can be learned, and @hurley-oldford suggest repeating variables to alleviate this. Another display that relies on variable ordering is the heatmap, which is widely used to visualize cluster structure in bioinformatics. @Wilkinson2009-pj provides a comprehensive history of this display in the social and natural sciences.

@@

The recently developed slice tour [@Laa2020-js] can be applied to any of the above tour types. It displays a slice through the orthogonal space, as opposed to a projection, and is explained in Section \@ref(slices).

@@

The smooth progression in the tour path is due to geodesic interpolation between the target planes. This takes into consideration two important aspects: (1) it maintains the orthonormality of the projection bases, (2) contains all the rotation to be between planes, not a particular basis in any plane. The first is clearly important because it ensures that we are looking at low-dimensional projections of the data always. The latter is harder to explain, but really important from a visual perspective. It stops any within-plane spin, and could be considered stabilizing the view. A more detailed explanation of this can be found in @Buja2005-cx.

@@

Various displays can be used to show projections in 1D, 2D and higher dimensions. A 1D projection displays the data in a histogram analogous to a shadow puppet being projected onto a wall, while 2D projections are displayed as scatterplots. Higher dimensional projections can be shown with multivariate displays like SPLOMs or parallel coordinates. Figure \ref{fig:display} shows a variety of displays for projections and these displays are useful to show the non-normal distributions in the projected data, clustering structure, and multivariate relationship between variables. It is possible to define tours for more complex data also. For example, @Martinez2013 proposed the image grand tour for multivariate spatial raster data, which displays a linear combination of variables at each spatial pixel.

@@

## Slices, sections and projections {#slices}

@@

There has been a long history of software including tour methods, or something similar. These have laid the foundation for the current tools. The American Statistical Association Statistical Graphics Video Library web site [@ASASGVL] hosts videos documenting the history. It is fabulous watching the videos titled "Multidimensional Scaling" (Kruskal, 1962), "Real-Time Rotation" (Chang, 1970) and "Prim-9" (Tukey, 1973) show some preliminary methods leading to the development of the grand tour. The video titled "Use of the Grand Tour in Remote Sensing" (McDonald and Willis, 1987) is the first to show a tour, and the video titled "Dataviewer: A Program for Looking at Data in Several Dimensions" (Buja and Tukey, 1987) demonstrates the tour as part of a larger data analysis system. The videos titled "XGobi: Dynamic Graphics for Data Analysis" (Swayne et al, 1991) and "Grand Tour and Projection Pursuit" (Cook et al, 1993) show the tour tools in the XGobi software system. Some important work on tours, for example @Wegman92 and @Tierney1991, @salch-scott and @Moustafa2010, is not documented in the video collection.
